Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)

### 2016-09-19 code/Rmd/CMC_ACC_StarAligned_TMM.Rmd
Added markdown to process STAR aligned ACC samples

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (DLPFC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_DLPFC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'DLPFC - STAR Reprocessed - with TMM')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (DLPFC)';

thisFileName <- 'CMC_DLPFC_StarAligned_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (DLPFC)
COUNT_ID <- 'syn7217916';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)
rownames(COUNT) = COUNT$geneID
COUNT$geneID = NULL
COUNT = data.matrix(COUNT)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                  "pH", "Dx", "CDR", "CauseOfDeath", "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                  "DLPFC_RNA_isolation: RNA Isolation Batch", "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?",
                  "DLPFC_RNA_report: Library Batch")

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RNA_Isolation_Batch,
                DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = DLPFC_RNA_isolation__RNA_Isolation_Batch,
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_PFC_260', 'MSSM_RNA_PFC_60', 'MSSM_RNA_PFC_139', 'MSSM_RNA_PFC_361', 
                           'MSSM_RNA_PFC_348', 'MSSM_RNA_PFC_389', 'PENN_RNA_PFC_56', 'PENN_RNA_PFC_59'))) %>%
  as.data.frame()
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape=Institution, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape = Institution, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "DLPFC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_DLPFC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_DLPFC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_DLPFC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_DLPFC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'CMC_DLPFC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_DLPFC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_DLPFC_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('CMC_DLPFC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'CMC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_DLPFC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
|  -------                  |   ---------                      |
|  Covariates               |  `r COV_OBJ$properties$id`       |
|  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
|  Design Matrix            |  `r DM_OBJ$properties$id`        |
|  Residual Expression      |  `r EXP_OBJ$properties$id`       |
|  Differential Expression  |  `r DEXP_OBJ$properties$id`      |

>>>>

---
title: "Covariate analysis of CMC RNASeq data reprocessed with Star-HTSeq pipeline (ACC with TMM normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(knitr)
library(knit2synapse)
library(synapseClient)

@@

knit2synapse::knitToFolderEntity(file = "./CMC_ACC_StarAligned_TMM.Rmd",
                                 parentId ="syn5569976",
                                 entityName = 'ACC - STAR Reprocessed - with TMM')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
# Synapse parameters
parentId = 'syn5569976';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of star aligned CMC reprocessed RNASeq data with TMM normalisation (ACC)';

thisFileName <- 'CMC_ACC_StarAligned_TMM.Rmd'

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ID <- 'syn7217917';
ALL_USED_IDs <- COUNT_ID
COUNT <- read.table(synGet(COUNT_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)
rownames(COUNT) = COUNT$geneID
COUNT$geneID = NULL
COUNT = data.matrix(COUNT)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Ethnicity", "Age of Death", "PMI (in hours)", 
                          "pH", "Dx", "CDR", "CauseOfDeath", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RNA Isolation Batch", "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?",
                          "ACC_RNA_report: Library Batch")

@@

METADATA = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Ethnicity, Age_of_Death, PMI__in_hours_, pH, Dx, CDR, CauseOfDeath,
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RNA_Isolation_Batch,
                ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_, ACC_RNA_report__Library_Batch) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                IsolationBatch = ACC_RNA_isolation__RNA_Isolation_Batch,
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch),
                IsolationBatch = as.factor(IsolationBatch))
```

### Data preprocessing
```{r preprocess.data, echo=TRUE}
# Remove samples marked as exclude and samples with no Gender, Ethnicity, PMI, and RIN
METADATA <- METADATA %>%
  filter(SampleID %in% colnames(COUNT)) %>%
  filter(IsolationExclude == 0, ReportExclude == 0) %>%
  filter(!is.na(PMI)) %>%
  filter(!is.na(Gender)) %>%
  filter(!is.na(Ethnicity)) %>%
  filter(!is.na(RIN)) %>%
  filter(!(SampleID %in% c('MSSM_RNA_ACC_250', 'MSSM_RNA_ACC_170'))) %>%
  as.data.frame()
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape=Institution, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Dx, shape = Institution, size=RIN))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:9)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p

@@

p = ggplot(DE, aes(y = -log10(adj.P.Val), x = logFC)) + geom_point() + xlim(c(-1,1))
p = p + geom_hline(yintercept = -log10(0.05), color = 'red') 
p = p + geom_vline(xintercept = log2(1.2), color = 'red') 
p = p + geom_vline(xintercept = -log2(1.2), color = 'red') 
p = p + facet_grid(.~Comparison, scales = 'fixed')
p
```
### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "ACC - STAR Reprocessed - with TMM", parentId = parentId)
CODE <- synStore(CODE)

# Store covariates
COVARIATES = rownameToFirstColumn(COVARIATES, 'SampleID')
write.table(COVARIATES, file = 'CMC_ACC_Covariates.tsv', sep = '\t', row.names=F, quote=F)
COV_OBJ = File('CMC_ACC_Covariates.tsv', name = 'Covariates', parentId = CODE$properties$id)
COV_OBJ = synStore(COV_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store logCPM
write.table(VOOM.ADJUSTED.GENE_EXPRESSION$E, file = 'CMC_ACC_logCPM.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_logCPM.tsv', name = 'Raw counts (logCPM)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                     executed = thisFile, activityDescription = activityDescription)

# Store design matrix
write.table(DM1$design, file = 'CMC_ACC_Design.tsv', sep = '\t', row.names=F, quote=F)
DM_OBJ = File('CMC_ACC_Design.tsv', name = 'Design Matrix', parentId = CODE$properties$id)
DM_OBJ = synStore(DM_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                  executed = thisFile, activityDescription = activityDescription)

# Store residual gene expression
write.table(RESIDUAL.GENE_EXPRESSION, file = 'CMC_ACC_Expression.tsv', sep = '\t', row.names=F, quote=F)
EXP_OBJ = File('CMC_ACC_Expression.tsv', name = 'Normalised Adjusted Residual Expression (Dx added)', parentId = CODE$properties$id)
EXP_OBJ = synStore(EXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
DEXP_OBJ = synStore(DEXP_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                    executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                     |
  |  -------                  |   ---------                      |
  |  Covariates               |  `r COV_OBJ$properties$id`       |
  |  Raw counts (log CPM)     |  `r COUNT_OBJ$properties$id`     |
  |  Design Matrix            |  `r DM_OBJ$properties$id`        |
  |  Residual Expression      |  `r EXP_OBJ$properties$id`       |
  |  Differential Expression  |  `r DEXP_OBJ$properties$id`      |
